Stateless Model Checking with Data-Race Preemption Points

### 2016-05-20 oopsla-16/response.txt

We agree with reviewer A's concern about possible bias in the set of
benchmarks. To address this, we will add some analysis of the types of bugs
found. For example, in the Iterative Context Bounding control experiment, 82
bugs were found with preemption bound 1, 15 with bound 2, and 2 with bound 3.
This is consistent with the distribution in [1] and shows our bugs are not
biased towards requiring high preemption bounds.

Our focus on maximizing completed state spaces is motivated by [2], in which
developers prefer compositional testing (manually restricting a test's scope
to complete faster), which suggests that full completions of partial state
spaces are easier for humans to interpret than partial completions of full
spaces. Our evaluation also shows that subset state spaces generally uncover
bugs faster (QS-Sync-Only vs SSS-MC-DPOR). However, your suggestion to
sometimes complete larger jobs first, thereby pruning any subsumed smaller
jobs, is an insightful optimization which we could implement in future work.

If the race detector produces a candidate where only one of two orders is
feasible, the MC will subsequently classify it as a false positive.

To address reviewer B, we choose new preemption points to add for each
data-race candidate, one for each racing instruction. This produces several
new jobs, each with a different combination of the new ones with existing
ones. For the sake of soundness, which assumes an infinite CPU budget, it
would suffice to randomly choose among new jobs until all are exhausted, or
even to always choose the biggest (maximal) job and ignore subset jobs.
Section 5's proofs assume this simple approach (call it Algorithm 0).
Algorithm 1 in section 3 optimizes algorithm 0 for finite CPU budgets to
prioritize finding bugs faster. Sorry for the confusion and we will clarify
this, including Algorithm 0 explicitly.

Regarding the suggestion of Maximal Causality Reduction [3], we were not able
to do a direct comparison because that paper has no open-source
implementation. We agree it is not necessarily orthogonal to our approach, but
could be combined in future work when an implementation becomes available.

We will clarify in our introduction that we assume sequentially-consistent
hardware, although in principle our work could be extended to weak memory
models using the techniques in [4].

Regarding Figure 3, the contrived program is just an example to explain the
background. We agree other approaches exist which can solve this case.

Reviewer C insightfully observes that Figure 7's count of bugs excludes
data-race candidates from the control experiments. This is the mentioned
philosophical difference about data races versus concrete failures; to address
this, we will add a table listing how many data races the control experiments
reported, and how many of those Quicksand refuted as false positives.

Reviewer D mentions that both HB and Limited HB will find all data races when
combined with model checking. We suspected this was true, but were only able
to prove the case for Limited HB. If you could provide a citation which proves
the case of regular HB combined with model checking, we are interested to read
it and would gladly include it for cam-ready. However, one further advantage
of Limited HB is that it reports candidates sooner, allowing Quicksand to test
them immediately, at the expense of more false positives. We will add a
discussion of this trade-off.

Regarding our malloc-recycle solution, we tried your suggestion to never
reallocate memory, but this unacceptably impacted the performance of some
malloc-heavy tests. Our proof allows eliminating these data-race candidates
without hacking the allocator.

Thanks for the clarification about FastTrack and CHESS; we will improve our
discussion of those tools.

Finally, reviewers A and D both criticized the formalism for unclear
definitions and notation. We agree and will work to improve the foundation and
notation definitions for cam-ready. For example, in the definition of MHB, we
define "cannot be reordered" in terms of the ability for threads to be
scheduled at prior preemption points, with no intermediate blocking operation
by the latter thread (this coincides with DPOR's approach to identify
alternate interleavings to explore).

[1] Musuvathi and Qadeer, Iterative Context Bounding, PLDI07
[2] Ball et al, Preemption Sealing, TACAS10
[3] Huang, Maximal Causality Reduction, PLDI15 ([24] in the paper/reviews)
[4] Zhang et al, DPOR for relaxed memory, PLDI15

### 2016-05-20 oopsla-16/response.txt

We agree with reviewer A about possible bias in the benchmarks. To address
this, we will add some analysis of the types of bugs found. For example, in
the Iterative Context Bounding control experiment, 82 bugs were found with
preemption bound 1, 15 with bound 2, and 2 with bound 3, consistent
with the distribution in [Musuvathi PLDI07]. This shows no bias towards
requiring high preemption bounds.

Maximizing completed state spaces is motivated by [Ball TACAS10], in which
developers prefer compositional testing (manually restricting a test's scope
to complete faster), which suggests that full completions of partial state
spaces are easier for humans to interpret than partial completions of full
spaces. We also showed subset state spaces generally uncover
bugs faster (QS-Sync-Only vs SSS-MC-DPOR). However, your suggestion to
sometimes complete larger jobs first, to prune subsumed smaller jobs, is an
insightful optimization for future work.

Reviewer B: We choose new preemption points to add for each
data-race candidate, one for each racing instruction. This produces several
new jobs, each with a different combination of the new ones with existing
ones. For the sake of soundness, which assumes an infinite CPU budget, it
would suffice to randomly choose among new jobs until all are exhausted, or
even to always choose the biggest (maximal) job and ignore subset jobs.
Section 5's proofs assume this simple approach (call it Algorithm 0).
Algorithm 1 in section 3 optimizes algorithm 0 for finite CPU budgets,
choosing the state space with smallest ETA. We will clarify this, including
Algorithm 0 explicitly.

Regarding the suggestion of Maximal Causality Reduction [Huang PLDI15], we
were not able to do a direct comparison because that paper has no open-source
implementation. This is a compelling direction for future work.

We will clarify in our introduction that we assume sequentially-consistent
hardware. Future work could extend our implementation and proofs using the
techniques in [Zhang PLDI15].

Figure 2 is meant to show the state of an arbitrary in-progress search,
although we agree it needs work. Regarding Figure 3, the contrived program is
just an example to explain the background. We agree other approaches exist
which can solve this case. Thanks for the suggestion about graph search; we
will look into it.

@@

Reviewer C insightfully observes that Figure 7's count of bugs excludes
data-race candidates from the control experiments. This is the mentioned
philosophical difference about data races versus concrete failures; to address
this, we will add a table listing how many data races the control experiments
reported, and how many of those Quicksand refuted as false positives.

Reviewer D mentions that both HB and Limited HB will find all data races when
combined with model checking. We suspected this was true, but have only proven
the case for Limited HB. Could you provide a citation for us to include during
revisions? One further advantage of Limited HB is reporting candidates sooner,
allowing Quicksand to test them immediately, at the expense of more false
positives. We will clarify this trade-off.

Regarding our malloc-recycle solution, we tried your never reallocating
memory, but this unacceptably impacted malloc-heavy tests. Our proof allows
eliminating these data-race candidates without hacking the allocator.

@@

Thanks for the clarification about FastTrack and CHESS; we will improve our
discussion of those tools.

Finally we agree with reviewers A and D and will work to improve the
formalism's notation and definitions. For example, in the definition of MHB,
we define "cannot be reordered" in terms of the latter thread's ability to be
scheduled at all intervening preemption points, with no intermediate blocking
operation (this coincides with DPOR's approach to identify alternate
interleavings to explore).

### 2016-07-14 oopsla-16/design.tex
refactor design section intro and write algorithm0

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening
%likewise
makes progressively deeper searches of the state space until the CPU budget is exhausted.
In this context, the depth is the number of PPs used.
Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs.
We refer to each unique set of PPs as a {\em job},
and prioritize them based on number of PPs, ETA, and whether they include data-race PPs,
We rely on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
We seek to maximize completed state spaces,
as each one serves as a guarantee that all interleavings possible with its PPs were tested.

@@

@@

\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}
With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When the MC reports an ETA too high for some job,
we compare with other pending and suspended jobs to find another one more likely to complete in time.
%
Our method,
%for doing so,
listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
%\footnote{
%Though its worst-case performance is $O(mn)$ in the
%%number of pending and suspended jobs,
%sizes of $\mathcal{P}$ and $\mathcal{S}$,
%in practice the non-constant portion beyond line 4 runs very infrequently
%and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
%reveals that it might finish in time after all.
Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.

>>>>

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening
%likewise
makes progressively deeper searches of the state space until the CPU budget is exhausted.
In this context, the depth is the number of PPs used.
Hence,
%our tool
\quicksand~schedules multiple MC instances in parallel to
%simultaneously
test many different subsets of the available PPs,
We refer to each unique set of PPs as a {\em job}.

@@

\cbstart
\subsection{Changing State Spaces}
\label{sec:algorithm0}

To introduce the Iterative Deepening algorithm,
we first show a simple approach for handling new PPs in the absence of any CPU budget restriction.
%show how we prioritize smaller state spaces based on ETAs.

{\bf Na\"ive approach.}
Given unlimited CPU time for testing,
we would always switch to the new maximal state space whenever adding a new PP.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known PPs will be duplicate work.
Algorithm~\ref{alg:algorithm0} demonstrates this na\"ive approach.
It would be seeded with the set of all statically-known synchronization API PPs,
and invoked whenever a new data race PP is added.
%
Our proofs in \sect{\ref{sec:totalverif}},
being concerned with the verification guarantee provided when \quicksand~completes within the CPU budget,
are based on this simple version of Iterative Deepening.
%

\newcommand\AllPPs{\ensuremath{\mathcal{A}}}
\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}

\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
	\Input{$j$, the currently-running job}
	%\Input{$eta$, $j_0$'s predicted completion time}
	\Input{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
	%\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
	%\Input{$T$, the remaining time in the CPU budget}
	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
		return NewJob(\AllPPs) // New maximal state space
	}{
		return $j$ // $j$ is still maximal
	}
	\caption{Na\"ive Iterative Deepening method}
	\label{alg:algorithm0}
\end{algorithm}

{\bf Prioritizing smaller jobs.}
However, in tests where full verification is not feasible,
focusing on the maximal state space alone is likely to be fruitless.
%
Hence, we prioritize all subset jobs
\cbend
based on number of PPs, ETA, and whether they include data-race PPs,
We rely on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
We seek to maximize completed state spaces,
as each one serves as a guarantee that all interleavings possible with its PPs were tested.
\cbstart
The next three subsections will show how we schedule these smaller jobs
based on their PP sets and ETAs.
\cbend

@@

With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When the MC reports an ETA too high for some job,
we compare with other pending and suspended jobs to find another one more likely to complete in time.
%
Our method,
%for doing so,
listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
%\footnote{
%Though its worst-case performance is $O(mn)$ in the
%%number of pending and suspended jobs,
%sizes of $\mathcal{P}$ and $\mathcal{S}$,
%in practice the non-constant portion beyond line 4 runs very infrequently
%and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
%reveals that it might finish in time after all.
Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.
